---
title: "Statatistics and Data Analysis for Financial Engineering - Chapter 8 Copulas"
output:
  github_document:
    pandoc_args: --webtex
always_allow_html: yes

---

```{r }
library(copula)
cop_t_dim3 = tCopula(dim=3, param=c(-0.6,0.75,0), dispstr="un", df=1)
set.seed(5640)
rand_t_cop = rCopula(n=500, copula=cop_t_dim3)
pairs(rand_t_cop)
cor(rand_t_cop)
```

### Problem1: \ 
(a) What copula model has been sampled? Give the correlation matrix? \


t-copula, degrees of freedom 1, correlation matrix is:

```{r }
cor = matrix(1:9, nrow = 3, ncol = 3)
cor[1,] = c(1, -0.6, 0.75)
cor[2,] = c(-0.6, 1, 0)
cor[3,] = c(0.75, 0, 1)
print(cor)
```

(b) What is the sample size? 500 \

### problem 2: scatterplot: \
(a) Components 2 and 3 are uncorrelated. Do they appear independent? Why?

A: The scatterplots would have been uniformly distributed if they where independent, which they are not.\ \

(b) Do you see signs of tail dependence? Where? \

A: The non-uniformity mentioned in (a) is that there are more data in the
corners, which shows that extreme values tend to occur together, although
because of the zero correlation, a positive extreme value of one component is
equally likely to be paired with a positive or negative extreme value of the other
component. \ \

(c) What are theeffects of dependence upon the plots? \

A: The effects of tail dependence is the tendency of extreme values to pair.
The negative correlation of components 1 and 2 shows in the concentration of
the data along the diagonal from upper left to lower right. Positive extreme
values in one component tend to pair with negative extreme values of ther other
component. \
The positive correlation of components 2 and 3 shows in the concentration of
the data along the diagonal from lower left to upper right. Positive extreme
values in one component tend to pair with positive extreme values of ther other
component \\

(d) The nonzero correlations in the copula do not have the same values as the corresponding sample correlations. Why? If it is not random variation, what is it? Hint: get confidence intervals for Pearson correlations with:
```{r }
cor.test(rand_t_cop[,1], rand_t_cop[,3])
```

which gives 95% CI for component 1 and 3. Does this CI include 0.75? \

A: The confidence interval is (0.6603, 0.7484) which
does not quite include 0.75. This is not surprising. 0.75 is the correlation
between the t-distributed random variables that define the copula and need not
be the same as the uniformly-distributed variables in the copula itself. \\

### problem 3: Simulation from copula models

1 Define a gauss-copula, 2. define multivariate distribution by specifying g-copula and marginal pdf. 3. generate random sample from the multivariate pdf.

```{r }
library(copula)
cop_t_dim3 = tCopula(dim = 3, param = c(-0.6,0.75,0), 
                     dispstr = "un", df = 1)
set.seed(5640)
rand_t_cop = rCopula(n = 500, copula = cop_t_dim3)
pairs(rand_t_cop)
cor(rand_t_cop)


cor.test(rand_t_cop[,1],rand_t_cop[,3])


cop_normal_dim3 = normalCopula(dim = 3, param = c(-0.6,0.75,0), 
                               dispstr = "un")
mvdc_normal = mvdc(copula = cop_normal_dim3, margins = rep("exp",3),
                   paramMargins = list(list(rate=2), list(rate=3), 
                                       list(rate=4)))
set.seed(5640)
rand_mvdc = rMvdc(n = 1000, mvdc = mvdc_normal)
pairs(rand_mvdc)
par(mfrow = c(2,2))
for(i in 1:3) plot(density(rand_mvdc[,i]))
for(i in 1:3) print(mean(rand_mvdc[,i]))
```

(a) What are the marginal distributions of the three components in rand_mvdc? What are their expected values?

A: They look Beta distributed. Expected values: 0.5, 0.34, 0.25

(b) Are the second and third components independent? Why?

A: The scatterplots would have been uniformly distributed if they where independent, which they are not.\ \

### Problem 4: Fitting Copula Models to Bivariate Return Data - IBM and S&P 500 - kendalls tau

```{r, message=FALSE, warning=FALSE  }
# Data download 9/9/2014
# library(quantmod)
# getSymbols(c("IBM", "^GSPC"), from="2004-06-01",to="2014-05-31") 
# IBM.SP500 = cbind(IBM[,6],GSPC[,6]) ; head(IBM.SP500)
# netReturns = ((diff(IBM.SP500)/lag(IBM.SP500)*100)[-1,]) ; tail(netReturns)
# colnames(netReturns) = c("IBM", "SP500") ; colnames(netReturns)
# head(netReturns) ; tail(netReturns)
# write.zoo(netReturns,"IBM_SP500_04_14_daily_netRtns.csv", index.name="Date", sep=",")

library(MASS)     #  for fitdistr() and kde2d() functions
library(copula)   #  for copula functions
library(fGarch)   #  for standardized t density
netRtns = read.csv("datasets/IBM_SP500_04_14_daily_netRtns.csv", header = T)
ibm = netRtns[,2]
sp500 = netRtns[,3]
est.ibm = as.numeric( fitdistr(ibm,"t")$estimate )
est.sp500 = as.numeric( fitdistr(sp500,"t")$estimate )
est.ibm[2] = est.ibm[2] * sqrt( est.ibm[3] / (est.ibm[3]-2) )
est.sp500[2] = est.sp500[2] * sqrt(est.sp500[3] / (est.sp500[3]-2) )


cor_tau = cor(ibm, sp500, method = "kendall")
print(cor_tau)
omega = sin((pi/2)*cor_tau) #0.5 ######### need to get correct value   

cop_t_dim2 = tCopula(omega, dim = 2, dispstr = "un", df = 4)

data1 = cbind(pstd(ibm, est.ibm[1], est.ibm[2], est.ibm[3]), 
              pstd(sp500, est.sp500[1], est.sp500[2], est.sp500[3]))
n = nrow(netRtns) ; n
data2 = cbind(rank(ibm)/(n+1), rank(sp500)/(n+1))
ft1 = fitCopula(cop_t_dim2, data1, method="ml", start=c(omega,4) ) 
ft2 = fitCopula(cop_t_dim2, data2, method="ml", start=c(omega,4) ) 
```

Q: Find value for Omega.

A: omega =  sin((pi/2)*cor_tau) = 0.7018346

### Problem 5 - Spearmans rank correlation

(a) Explain the difference between ft1 and ft2

A: Both fits are by pseudo-likelihood. ft1 is the parametric approach because
the univariate marginal distributions are estimated by fitting t-distributions,
and ft2 is the nonparametric approach because the univariate distributions are
estimated by empirical CDFs.

(b) Do the two estimates seeem significatly different?

A The two estimates of the correlation are 0.7022 and 0.7031. The two estimates of the degrees of freedom are 2.98 and 3.02. Thus, the two estimates of
the copula are quite similar with no significant practical difference. Notice also
that the two estimates of the correlation are similar to the estimate, 0.7018, in
Problem 3 that used Kendall’s tau

### problem 6: 

```{r}
mvdc_t_t = mvdc( cop_t_dim2, c("std","std"), list(
           list(mean=est.ibm[1],sd=est.ibm[2],nu=est.ibm[3]),
           list(mean=est.sp500[1],sd=est.sp500[2],nu=est.sp500[3])))


# Will run for 1. minute or more
#fit_cop = fitMvdc(cbind(ibm,sp500),mvdc_t_t,start=c(ft1@estimate,est.ibm,est.sp500), hideWarnings=FALSE)
#print(fit_cop)

start = c(est.ibm, est.sp500, ft1@estimate)
objFn = function(param) -loglikMvdc(param,cbind(ibm,sp500),mvdc_t_t)
tic = proc.time()
ft = optim(start, objFn, method="L-BFGS-B",
           lower = c(-.1,0.001,2.2, -0.1,0.001,2.2,  0.2,2.5),
           upper = c( .1,   10, 15,  0.1,   10, 15,  0.9, 15) )
toc = proc.time()
total_time = toc - tic ; total_time[3]/60
print(total_time)
print(ft)
```

Lower and upper bounds are used to constrain the algorithm to search where log-likeliehood is defined and finite.
FitMvdc() does not allow to define lower and upper bound and did not converge on this problem. 

(a) What are the estimates of the copula parameters in fit_cop? 

A:
```{=html}
> fit_cop = fitMvdc(cbind(ibm,sp500),mvdc_t_t,start=c(ft1@estimate,est.ibm,est.sp500), hideWarnings=FALSE) 
Error in optim(start, loglikMvdc, mvdc = mvdc, x = data, method = method,  : 
  initial value in 'vmmin' is not finite
In addition: Warning messages:
1: In sqrt(nu/(nu - 2)) : NaNs produced
2: In sqrt(nu/(nu - 2)) : NaNs produced
3: In sqrt(nu/(nu - 2)) : NaNs produced
4: In sqrt(nu/(nu - 2)) : NaNs produced
```

(b) What are the estimates of the parameters in the univariate marginal distribution? 

A:

(c) Was the estimation method maximum likelihood, semiparametric pseudo ml, or parametric pseudo-ml?

(d) Estimate the coefficient of lower tail dependence for this copula.

### Problem 7:

```{r}
fnorm = fitCopula(copula = normalCopula(dim=2), data=data1, method="ml" ) 
ffrank = fitCopula(copula = frankCopula(3, dim = 2), data = data1, method = "ml" )
fclayton = fitCopula(copula = claytonCopula(1, dim=2), data = data1, method = "ml" ) 
fgumbel = fitCopula(copula = gumbelCopula(3, dim=2), data = data1, method = "ml" ) 
fjoe = fitCopula(copula=joeCopula(2,dim=2),data=data1,method="ml" ) 
```

The estimated copulas (CDFs) will be compared with the empirical copulas

```{r, warning=FALSE}
Udex = (1:n)/(n+1)

#Cn: The Empirical Copula
tmp_u = cbind(rep(Udex,n),rep(Udex,each=n))
#Cn = C.n(u=tmp_u, U=data1, method="C")
Cn = C.n(tmp_u, data1)

EmpCop = expression(contour(Udex, Udex, matrix(Cn, n, n), col = 2, add = TRUE))
par(mfrow=c(2,3))
contour(tCopula(param=ft$par[7],dim=2,df=round(ft$par[8])), 
        pCopula, main = expression(hat(C)[t]), mgp = c(2.5,1,0), 
        xlab = expression(hat(U)[1]), ylab = expression(hat(U)[2]) )
eval(EmpCop)
contour(normalCopula(param=fnorm@estimate[1], dim = 2), 
        pCopula, main = expression(hat(C)[Gauss]), mgp = c(2.5,1,0),
        xlab = expression(hat(U)[1]), ylab = expression(hat(U)[2]) )
eval(EmpCop)
contour(frankCopula(param=ffrank@estimate[1], dim = 2), 
        pCopula, main = expression(hat(C)[Fr]), mgp = c(2.5,1,0),
        xlab = expression(hat(U)[1]), ylab = expression(hat(U)[2]) )
eval(EmpCop)
contour(claytonCopula(param=fclayton@estimate[1], dim = 2), 
        pCopula, main = expression(hat(C)[Cl]), mgp = c(2.5,1,0),
        xlab = expression(hat(U)[1]), ylab = expression(hat(U)[2]) )
eval(EmpCop)
contour(gumbelCopula(param=fgumbel@estimate[1], dim = 2), 
        pCopula, main = expression(hat(C)[Gu]), mgp = c(2.5,1,0),
        xlab = expression(hat(U)[1]), ylab = expression(hat(U)[2]) )
eval(EmpCop)
contour(joeCopula(param=fjoe@estimate[1], dim = 2), 
        pCopula, main = expression(hat(C)[Joe]), mgp = c(2.5,1,0),
        xlab = expression(hat(U)[1]), ylab = expression(hat(U)[2]) )
eval(EmpCop)

```

### Problem 8

A 2-d KDE of the copulas density will be compared with the parametric density estimate (PDFs)
```{r}
par(mfrow=c(2,3))
contour(tCopula(param=ft$par[7],dim=2,df=round(ft$par[8])),
        dCopula, main = expression(hat(c)[t]), mgp = c(2.5,1,0), 
  nlevels=25, xlab=expression(hat(U)[1]),ylab=expression(hat(U)[2]))
contour(kde2d(data1[,1],data1[,2]), col = 2, add = TRUE)
contour(normalCopula(param=fnorm@estimate[1], dim = 2), 
        dCopula, main = expression(hat(c)[Gauss]), mgp = c(2.5,1,0),
  nlevels=25, xlab=expression(hat(U)[1]),ylab=expression(hat(U)[2]))
contour(kde2d(data1[,1],data1[,2]), col = 2, add = TRUE)
contour(frankCopula(param=ffrank@estimate[1], dim = 2), 
        dCopula, main = expression(hat(c)[Fr]), mgp = c(2.5,1,0),
  nlevels=25, xlab=expression(hat(U)[1]),ylab=expression(hat(U)[2]))
contour(kde2d(data1[,1],data1[,2]), col = 2, add = TRUE)
contour(claytonCopula(param=fclayton@estimate[1], dim = 2), 
        dCopula, main = expression(hat(c)[Cl]), mgp = c(2.5,1,0),
  nlevels=25, xlab=expression(hat(U)[1]),ylab=expression(hat(U)[2]))
contour(kde2d(data1[,1],data1[,2]), col = 2, add = TRUE)
contour(gumbelCopula(param=fgumbel@estimate[1], dim = 2), 
        dCopula, main = expression(hat(c)[Gu]), mgp = c(2.5,1,0),
  nlevels=25, xlab=expression(hat(U)[1]),ylab=expression(hat(U)[2]))
contour(kde2d(data1[,1],data1[,2]), col = 2, add = TRUE)
contour(joeCopula(param=fjoe@estimate[1], dim = 2), 
        dCopula, main = expression(hat(c)[Joe]), mgp = c(2.5,1,0),
  nlevels=25, xlab=expression(hat(U)[1]),ylab=expression(hat(U)[2]))
contour(kde2d(data1[,1],data1[,2]), col = 2, add = TRUE)


```

``` {r}
```