---
title: "Statatistics and Data Analysis for Financial Engineering - Chapter 8 Copulas"
output:
  github_document:
    pandoc_args: --webtex
always_allow_html: yes

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- https://latex.codecogs.com/svg.latex? -->

A copula is a multivariate CDF whose univariate marginal distribution are all U(0,1).

By Sklar's theorem: 
$$
F_Y(y_1,..,y_d) = C_Y(F_{Y_1}(y_1),..,F_{Y_d}(y_d))
$$

data and code from the book can be found here: https://people.orie.cornell.edu/davidr/SDAFE2/

## Special copulas

1. d-dimensional independent copula $C_0$ of $U(0,1)^d$ \
2. co-monotonicity $C_+$ describes positive dependence and $C_+(u_1,..,u_d) = min(u1,..,u_d)$ \
3. counter-monotonicity copula $C_-$ which has negative dependence and $C_-(u_1, u_2) = max(u_1, u_2) if d \leq 2$.
If d > 2 then a lower bound for copulas: $max(u_1+..+u_d+1-d, 0) \leq c(u_1,..,u_d)$ \

For Gaussian and t-copulas, let $\Omega$ be the correlation matrix: \
1. $\Omega \rightarrow C_y$ is 1-1 to the Gauss copula   \
2. $\Omega = \textbf{I} \rightarrow$ Meta-Gaussian distribution, which is the independent copula (1.) \
3. $\Omega = \textbf{1} \rightarrow C_+$ \
4. $\Omega = \textbf{-1} \rightarrow C_-$ \


## Archimedean Copulas

Archimedean copula with generator function:


$$
C(u_1,..,u_d) = \varphi^{-1}(\varphi(u_1)+..+\varphi(u_d))
$$

Satisfies 3 axioms: \
1. $\varphi$ is continuous, strictly increasing, and convex on $\varphi : [0,1] \rightarrow [0, \infty)$ \
2. $\varphi(0) = \infty$ \
3. $\varphi(1) = 1$ \

#### Frank copula

Generator function:
$$
C(u|\theta) = -ln(\frac{e^{-\theta u} -1)}{e^{-\theta} -1)}), -\infty < \theta < \infty
$$
$\theta \rightarrow 0 \implies c(u_1, u_2) \rightarrow C_0$ \
$\theta \rightarrow \infty \implies c(u_1, u_2) \rightarrow C_+$ \
$\theta \rightarrow -\infty \implies c(u_1, u_2) \rightarrow C_-$ \

#### Crayton copula

Generator function:
$$
C(u|\theta) = \frac{1}{\theta}(u^{-\theta}-1), \theta > 0
$$
$\theta \rightarrow 0 \implies c(u) \rightarrow C_0$ \
$\theta \rightarrow \infty \implies c(u_1, u_2) \rightarrow C_+$ \
$\theta \rightarrow -1 \implies c(u_1, u_2) \rightarrow C_-$ \


#### Gumbel Copula adsfads

Generator function:
$$
C(u|\theta) = -ln(u)^{\theta}, \theta \geq 1
$$

$\theta = 1 \implies c(u) \rightarrow C_0$ \
$\theta \rightarrow \infty \implies c(u_1, u_2) \rightarrow C_+$ \




#### Joe copula

Generator function:
$$
C(u|\theta) = -ln(1-(1-\theta)^{\theta}), \theta \geq 1
$$

$\theta =1  \implies c(u) \rightarrow C_0$ \
$\theta \rightarrow \infty \implies c(u_1, u_2) \rightarrow C_+$ \

Joe copula is similar to Gumbel. It cannot have negative dependence. It allows stronger upper tail
dependence and is closer to being a reverse Clayton copula in the positive dependence case.
is closer to a reverse Clayton copula.

### Plot of generator function for Frank cupola

```{r Frank Copula generator function}
library(copula)
u= seq(0.000001, 1, length=500)
frank = iPsi(copula=archmCopula(family="frank", param=1), u)
plot(u, frank, type="l", lwd=3, ylab=expression(phi(u)))
abline(h=0)
abline(v=0)
```

### Scatter plot of 9 bivariate Frank copulas

```{r Scatterplot of 9 frank copulas, message=FALSE}
set.seed(5640)
theta = c(-100, -50, -10, -1, 0, 5, 20, 50, 500)
par(mfrow=c(3,3), cex.axis=1.2, cex.lab=1.2, cex.main=1.2)
for(i in 1:9){
  U= rCopula(n=200, copula=archmCopula(family="frank", param=theta[i]))
  plot(U, xlab=expression(u[1]), ylab=expression(u[2]), 
                          main=eval(substitute(expression(paste(theta, " = ", j)),
                         list(j = as.character(theta[i])))))
}
```
_Figure 2: Bivariate random samples of size 200 from various Frank copulas._

### Scatter plot of 9 bivariate Clayton copulas
```{r Scatterplot of 9 clayton copulas, echo=FALSE}
set.seed(5640)
theta = c(-0.98, -0.7, -0.3, -0.1, 0.1, 1, 5, 15, 100)
par(mfrow=c(3,3), cex.axis=1.2, cex.lab=1.2, cex.main=1.2)
for(i in 1:9){
  U= rCopula(n=200, copula=archmCopula(family="clayton", param=theta[i]))
  plot(U, xlab=expression(u[1]), ylab=expression(u[2]), 
                          main=eval(substitute(expression(paste(theta, " = ", j)),
                         list(j = as.character(theta[i])))))
}
```
_Figure 3: Bivariate random samples of size 200 from various clayton copulas.-



### Scatter plot of 6 bivariate Gumbel copulas
```{r Scatterplot of 9 gumbel copulas, echo=FALSE}
set.seed(5640)
theta = c(1.1, 1.5, 2, 4, 8, 50)
par(mfrow=c(3,3), cex.axis=1.2, cex.lab=1.2, cex.main=1.2)
for(i in 1:6){
  U= rCopula(n=200, copula=archmCopula(family="gumbel", param=theta[i]))
  plot(U, xlab=expression(u[1]), ylab=expression(u[2]), 
                          main=eval(substitute(expression(paste(theta, " = ", j)),
                         list(j = as.character(theta[i])))))
}
```
_Figure 4: Bivariate random samples of size 200 from various gumbel copulas._

### Scatter plot of 6 bivariate Joe copulas
```{r Scatterplot of 9 joe copulas, echo=FALSE}
set.seed(5640)
theta = c(1.1, 1.5, 2, 4, 8, 50)
par(mfrow=c(3,3), cex.axis=1.2, cex.lab=1.2, cex.main=1.2)
for(i in 1:6){
  U= rCopula(n=200, copula=archmCopula(family="joe", param=theta[i]))
  plot(U, xlab=expression(u[1]), ylab=expression(u[2]), 
                          main=eval(substitute(expression(paste(theta, " = ", j)),
                         list(j = as.character(theta[i])))))
}
```
_Figure 5: Bivariate random samples of size 200 from various joe copulas._

## Rank correlation: 1. Kendall's Tau, 2. Spearman's rank correlation coefficient

## Tail dependence

```{r Scatterplot  }
rho = seq(-1, 1, by=0.01)
df = c(1,4,25, 240)
x1 = -sqrt((df[1]+1)*(1-rho)/(1+rho))
lambda1 = 2*pt(x1, df[1]+1)
x4 = -sqrt((df[2]+1)*(1-rho)/(1+rho))
lambda4 = 2*pt(x4, df[2]+1)
x25 = -sqrt((df[3]+1)*(1-rho)/(1+rho))
lambda25 = 2*pt(x25, df[3]+1)
x250 = -sqrt((df[4]+1)*(1-rho)/(1+rho))
lambda250 = 2*pt(x250, df[4]+1)
par(mfrow=c(1,1), lwd=2, cex.axis=1.2, cex.lab=1.2)
plot(rho, lambda1, type="l", lty=1, xlab=expression(rho), ylab=expression(lambda[l]==lambda[u]))
lines(rho, lambda4, lty=2)
lines(rho, lambda25, lty=2)
lines(rho, lambda250, lty=2)
legend("topleft", c( expression(nu==1), expression(nu=4), expression(nu==25), expression(nu=250) ), lty=1:4)


```


```{r example232 }
getwd()
```

## Example: flows in pipeline

```{r example }
library(copula)
library(sn)
dat = read.csv("datasets/FlowData.csv")
dat = dat/10000
n = nrow(dat)
x1 = dat$Flow1
fit1 = st.mple(matrix(1,n,1), y=x1, dp=c(mean(x1), sd(x1),0,10))
est1 = fit1$dp              #vector or list of estimated DP parameters
u1 = pst(x1, dp=est1)       # vector of probabililities - skew-t
x2 = dat$Flow2
fit2 = st.mple(matrix(1,n,1), y=x2, dp=c(mean(x2), sd(x2),0,10))
est2 = fit2$dp
u2 = pst(x2, dp=est2)
U.hat = cbind(u1,u2)
z1 = qnorm(u1)
z2 = qnorm(u2)
Z.hat = cbind(z1,z2)
```

```{r example plot}
library(ks) 
fhatU = kde(x=U.hat, H=Hscv(x=U.hat))
par(mfrow=c(2,2), cex.axis=1.2, cex.lab=1.2, cex.max=1.2)
hist(u1, main="(a)", xlab=expression(hat(U)[1]), freq=FALSE)
hist(u1, main="(b)", xlab=expression(hat(U)[2]), freq=FALSE)
plot(u1, u2, main="(c)", xlab = expression(hat(U)[1]), ylab = expression(hat(U)[2]), mgp = c(2.5, 1, 0))
plot(fhatU, drawpoints=FALSE, drawlabels=FALSE, cont=seq(10, 80, 10), 
     main="(d)", xlab=expression(hat(U)[1]), ylab=expression(hat(U)[2]), mgp = c(2.5, 1, 0)) 

```
-Figure 6: Pipeline data. Density histograms (a), and (b) and a scatterplot (c) of the 
uniform-transformed flows. The empirical copula C.hat, is the empirical CDF of the  data in (c).
Contours (d) from an estimated copula density c.hat via a two-dimensional KDE of (c)_

```{r example plot2}

fhatZ = kde(x=Z.hat, H=Hscv(x=Z.hat))

#pdf("norm_flows_hist_plot.pdf", width=7, height=7)
#
par(mfrow=c(2,2), cex.axis=1.2, cex.lab=1.2, cex.main=1.2)
qqnorm(z1, datax=T, main="(a)") ; qqline(z1)
qqnorm(z2, datax=T, main="(b)") ; qqline(z2)
plot(z1, z2, main="(c)", xlab = expression(hat(Z)[1]), ylab = expression(hat(Z)[2]), mgp = c(2.5, 1, 0))
plot(fhatZ, drawpoints=FALSE, drawlabels=FALSE, cont=seq(10, 90, 10), 
     main="(d)", xlab=expression(hat(Z)[1]), ylab=expression(hat(Z)[2]), mgp = c(2.5, 1, 0)) 
```
_Figure 7: Pipeline date. Normal quantile plots (a) and (b), a scatterplot (c) and KDE density countours 
from the normal-transformed flows._

```{r example table}
options(digits=3)
cor.test(u1, u2, method="spearman")
cor.test(u1, u2, method="kendall")
sin(-0.242*pi/2)
cor.test(u1, u2, method="pearson")
cor.test(z1, z2, method="pearson")
```
_Table 1: Estimates of copula parameters, maximized log-likelihood, and AIC using the uniform-transfomred pipline 
flow data._


Next: fitting the paramteric pseudo-maximum likelihood: Btw how lazy is this code in the text-book! .Last.value is such a hack. 
```{r example table2}
library(knitr)
library(kableExtra)
omega = -0.371

options(digits=4)

Ct = fitCopula(copula=tCopula(dim = 2), data=U.hat, method="ml", start=c(omega, 10)) 
#Ct@estimate
mle_t = loglikCopula(param=Ct@estimate, U.hat, copula=tCopula(dim = 2))
t_AIC = -2*mle_t + 2*length(Ct@estimate)
#
Cgauss = fitCopula(copula=normalCopula(dim = 2), data=U.hat, method="ml", start=c(omega)) 
#Cgauss@estimate
mle_g = loglikCopula(param=Cgauss@estimate, U.hat, copula=normalCopula(dim = 2))
g_AIC = -2*mle_g + 2*length(Cgauss@estimate)
# Not run
Cgu = fitCopula(copula=gumbelCopula(2, dim=2), data=U.hat, method="ml")
# Not run
Cjoe = fitCopula(copula=joeCopula(2, dim=2), data=U.hat, method="ml")
#
Cfr = fitCopula(copula=frankCopula(1, dim=2), data=U.hat, method="ml")
#Cfr@estimate
mle_f = loglikCopula(param=Cfr@estimate, U.hat, copula=frankCopula(dim = 2))
f_AIC = -2*mle_f + 2*length(Cfr@estimate)
#
Ccl = fitCopula(copula=claytonCopula(1, dim=2), data=U.hat, method="ml")
#Ccl@estimate
mle_c = loglikCopula(param=Ccl@estimate, U.hat, copula=claytonCopula(dim = 2))
c_AIC = -2*mle_c + 2*length(Ccl@estimate)

# Put the data in a dataframe:
df <- data.frame(CopulaFamily=c('t', '', 'Gaussian', 'Frank', 'Clayton'),
        variable=c("t.hat", "v.hat", "p.hat", "theta.hat","theta.hat"),
        estimates=c(Ct@estimate[1], Ct@estimate[2], Cgauss@estimate,  Cfr@estimate,Ccl@estimate[1]),
         maximized_ll=c(mle_t, "", mle_g, mle_f, mle_c),
         AIC=c(t_AIC, "", g_AIC, f_AIC, c_AIC))

df$maximized_ll <- as.numeric(df$maximized_ll)
df$AIC <- as.numeric(df$AIC)
kbl(df, digits=2)
```


